---
title: 'R-intro - Session 2 exercise solutions'
author: "João Gonçalves"
date: "19 de Julho de 2018"
output:
  pdf_document:
    toc: true
    toc_depth: 4
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Missing data in R (NA values)

---------------------------------------------------------------------------------------------------------

### Exercise 1

1a)

```{r}

avg_temp <- c(10.5, NA, 12.3, 14.4, 15.6, 17.7, NA, 20.4, 20.3, 17.1, 13.3, 10.5)

# Calculate the sum of the logical vector to count the number of NA's
sum(is.na(avg_temp))

```

1b)

```{r}

# When NA values exist in a vector to calculate the mean or the std-dev the option 
# na.rm must be set to TRUE

mean(avg_temp, na.rm= TRUE)

sd(avg_temp, na.rm= TRUE)

```

1c)

```{r}

# Don't forget to turn the option na.rm equal to TRUE to calculate the median
avg_temp[is.na(avg_temp)] <- median(avg_temp, na.rm = TRUE)

print(avg_temp)
```

1d)

```{r}

# Redefine the avg-temp again with NA values
avg_temp <- c(10.5, NA, 12.3, 14.4, 15.6, 17.7, NA, 20.4, 20.3, 17.1, 13.3, 10.5)


# Start by creating a vector for all vector indices
ind <- (1:length(avg_temp))

# Use vector ind to get the indices which are NA
ind_NA <- ind[is.na(avg_temp)]

# This is equal to using the which() function
ind_NA <- which(is.na(avg_temp))

# Replace values using vector positions
avg_temp[2] <- mean(avg_temp[c(1,3)])

avg_temp[7] <- mean(avg_temp[c(6,8)])


# More general solution using the ind_NA vector with NA indices
# Replace the first NA
avg_temp[ind_NA[1]] <- mean(avg_temp[c(ind_NA[1] - 1, ind_NA[1] + 1)])

# And the second NA
avg_temp[ind_NA[2]] <- mean(avg_temp[c(ind_NA[2] - 1, ind_NA[2] + 1)])

print(avg_temp)

```



# Dataframes in R

---------------------------------------------------------------------------------------------------------


### Quick exercise 1

QE1)

```{r}

precip <- data.frame(
  
  cities = c("Mobile", "Juneau", "Phoenix", "Little Rock", "Los Angeles", 
             "Sacramento", "San Francisco", "Denver", "Hartford", "Wilmington"),
  
  precipitation = c( 67.0, 54.7, 7.0, 48.5, 14.0, 17.2, 20.7, 13.0, 43.4, 40.2), 
  
  stringsAsFactors = FALSE
)

print(precip)

precip$cities

class(precip$cities)

```


# Exploratory data analysis - scatterplots and correlations

---------------------------------------------------------------------------------------------------------


### Exercise 2

2.

2a)

```{r}

# Number of rows with missing values
sum(!complete.cases(airquality))

# Clean dataframe using complete.cases() function
aq <- airquality[complete.cases(airquality), ]

# This is equal to using the na.omit() function to suppress rows with NA's
aq <- na.omit(airquality)

# Check which rows (by position) were omitted when function na.omit was used
attr(aq,"na.action")

```

2b)

```{r}

summary(aq)

```

2c)

```{r}

# Use plot and specifically declare the x and y vector
plot(x = aq$Wind, y = aq$Ozone)

# Or you can use the formula interface to get a similar effect
plot(Ozone ~ Wind, data = aq)

# EXTRA: add a regression line to the plot based on the abline() function
# Notice that a linear model (lm) is used as an input in abline
abline(lm(Ozone ~ Wind, data = aq), col="red", lwd=2)

# The correct option is...
print("iv")
```

2d)

```{r}

plot(aq)

# By visual inspection it seems that Temp (temperature), Wind (wind speed) and 
# Solar.R (solar radiation) are more clearly correlated with Ozone concentration

```


2e)

```{r}

# (i)
cor.test(aq$Ozone, aq$Solar.R, method="spearman")
## Rho = 0.3481865 p < 0.001 

# (ii)
cor.test(aq$Ozone, aq$Wind, method="spearman")
## Rho = -0.6051364 p < 0.001 

# (iii)
cor.test(aq$Ozone, aq$Temp, method="spearman")
## Rho = 0.7729319 p < 0.001 

```


## Dataframe indexation with [ , ]

 
### Exercise 3

3. 

3a)

```{r eval=FALSE}

aq[,-c(3:ncol(aq))]

```

3b)

```{r eval=FALSE}

aq[,c("Day","Month")]

```

3c)

```{r}

aq[1:3, ]

```

3d)

```{r}

nr <- nrow(aq)
aq[(nr-10):nr, ]

# Or more simply
tail(aq, n = 10)

```

3e)

```{r}

aq[aq$Ozone > mean(aq$Ozone), ]

```

3f)

```{r}

aq[aq$Ozone < quantile(aq$Ozone, probs=0.1), ]

```



## Slicing with `subset()`


### Quick exercise 2

QE2)

```{r}

# Combine each condition using the & (AND) operator

subset(aq, 
       Ozone > 80 & 
       Temp > 90 & 
       Wind < 5)

```



# Factor variables

---------------------------------------------------------------------------------------------------------



### Exercise 4

4.

4a)

```{r}

# Use the line below to call the data
data(iris)
head(iris) # This is used to make a first check of the data

# Check object class
class(iris$Species)

# Check if it is a factor variable
is.factor(iris$Species)

```

4b)

```{r}

levels(iris$Species)

```

4c)

```{r}

table(iris$Species)

```

4d)

```{r}

# The function by() can be used for calculating a function for each level in INDICES
by(iris$Petal.Length, INDICES = iris$Species, FUN = mean)

# Or using indexation the mean can be calculated this way for each species
mean(iris$Petal.Length[iris$Species == "setosa"])
mean(iris$Petal.Length[iris$Species == "versicolor"])
mean(iris$Petal.Length[iris$Species == "virginica"])
```

4e)

```{r}

# The function by() can be used for calculating a function for each level in INDICES
by(iris$Petal.Length, INDICES = iris$Species, FUN = sd)

# Or using indexation the std-dev can be calculated this way for each species
sd(iris$Petal.Length[iris$Species == "setosa"])
sd(iris$Petal.Length[iris$Species == "versicolor"])
sd(iris$Petal.Length[iris$Species == "virginica"])

```


# Analyzing distributions

---------------------------------------------------------------------------------------------------------


### Exercise 5

5.

5a)

```{r}

# This command will allow to make a 2 x 2 plot i.e. with the 4 boxplots 
par(mfrow=c(2,2))


# Now let's make the boxplots
# In boxplot it's possible to modify the title (main), 
# the y-axis label (ylab) and x-axis label (xlab)

boxplot(Sepal.Length ~ Species, data = iris, main = "Sepal length by Iris species", 
        ylab = "Sepal length (cm)", xlab = "Species")

boxplot(Sepal.Width ~ Species, data = iris, main = "Sepal width by Iris species", 
        ylab = "Sepal width (cm)", xlab = "Species")

boxplot(Petal.Length ~ Species, data = iris, main = "Petal length by Iris species", 
        ylab = "Petal length (cm)", xlab = "Species")

boxplot(Petal.Width ~ Species, data = iris, main = "Petal width by Iris species", 
        ylab = "Petal width (cm)", xlab = "Species")

```


5b)

i) Overall, petal dimensions seems to provide better differentiation between species with lower overlap between boxes/distributions.

ii) Both petal length and width provide very good differentiation.

iii) Sepal width seems to perform worst due to the higher overlap between boxes/distributions.



## Hypothesis testing


### Exercise 6


6.

6a)

```{r}

boxplot(weight ~ group, data = PlantGrowth)

```

6b)

```{r}

# Make a variable with weight measurements for the control group
weight_control <- PlantGrowth$weight[PlantGrowth$group == 'ctrl']

# Make a variable with weight measurements for the treatment-1 group
weight_trt1 <- PlantGrowth$weight[PlantGrowth$group == 'trt1']

# Now the t.test
t.test(x = weight_control, y = weight_trt1)

```

According to the t-Test there is not sufficient evidence to reject the null-hypothesis, H0: Avg(control) = Avg(trt1), given that the p-value > 0.05

6c)

```{r}

# Make a variable with weight measurements for the control group
weight_control <- PlantGrowth$weight[PlantGrowth$group == 'ctrl']

# Make a variable with weight measurements for the treatment-2 group
weight_trt2 <- PlantGrowth$weight[PlantGrowth$group == 'trt2']

# Now the t.test
t.test(x = weight_control, y = weight_trt2)

```

According to the t-Test there is sufficient evidence to reject the null-hypothesis, H0: Avg(control) = Avg(trt2), given that the p-value < 0.05



## Analysis of variance (one-way)



### Exercise 7

7.

7a) 

```{r}

aov_iris <- aov(Sepal.Width ~ Species, data = iris)

summary(aov_iris)

```

According to the ANOVA test there is sufficient evidence to reject the null-hypothesis, H0: Avg(setosa) = Avg(versicolor) = Avg(virginica), given that the p-value < 0.05

7b)

```{r}

aov_iris <- aov(Petal.Width ~ Species, data = iris)

summary(aov_iris)

```

According to the ANOVA test there is sufficient evidence to reject the null-hypothesis, H0: Avg(setosa) = Avg(versicolor) = Avg(virginica), given that the p-value < 0.05



### Exercise 8

8.

8a)

```{r}

# Read the csv data file 
# NOTE: depending on the working directory this path may not work and in that case you have to adjust it
diet <- read.csv("../Session-02/diet.csv")

```

8b)

```{r}

# Use na.omit() to quickly exclude rows with NA values 
diet <- na.omit(diet)

```

8c)

```{r}

# To add a new column to the dataset we can use either the function data.frame() or cbind()
diet <- data.frame(diet, Treatment = factor(diet$Diet, levels = c(1,2,3), 
                                            labels = c("Trt-A", "Trt-B", "Trt-C")))

# Solution with cbind()
# diet <- cbind(diet, Treatment = factor(diet$Diet, levels = c(1,2,3), labels = c("Trt-A", "Trt-B", "Trt-C")))

```

8d)

```{r}

diet <- data.frame(diet, Weight.Loss = diet$pre.weight - diet$weight6weeks)

#print(diet)

```

8e)

```{r}

aov_diet <- aov(Weight.Loss ~ Treatment, data = diet)

summary(aov_diet)

```

According to the ANOVA test there is sufficient evidence to reject the null-hypothesis, H0: Avg(Trt-A) = Avg(Trt-B) = Avg(Trt-C), given that the p-value < 0.05








