---
title: "Session 01 - Exercise solutions"
author: "João Gonçalves"
date: "14 de Julho de 2018"
output:
  html_document: default
  pdf_document:
    toc: true
    toc_depth: 4
---


# Using the help system

---------------------------------------------------------------------------------------------------------


### Exercise 1

```{r eval=FALSE, include=TRUE}

#1) 

# Mean function
?mean

# standard-deviation
?sd

# Calculate quantiles
?quantile

# Calculate the Median Absolute Deviation (MAD)
?mad

```


### Exercise 2

2a) 
```{r eval=FALSE, include=TRUE}

# 2a)

??'Weighted Arithmetic Mean'
help.search("Weighted Arithmetic Mean")

```

2b) 
```{r eval=FALSE, include=TRUE}

# 2b)
??'correlation'
help.search("correlation")

```

2c) 
```{r eval=FALSE, include=TRUE}

# 2c)
??'linear model'
help.search("linear model")

```


# Operations with vectors

---------------------------------------------------------------------------------------------------------


## Basic operations


### Exercise 3

Define the variables first:

```{r  eval=TRUE, include=TRUE}

x <- 10.2
y <- 5.7

```

3a) 
```{r  eval=TRUE, include=TRUE}

# 3a)
(x + y) / 2

```

3b) 
```{r  eval=TRUE, include=TRUE}

# 3b)
(1/2) * x^2 - log10(y)

```

3c)
```{r  eval=TRUE, include=TRUE}

# 3c)
z <- sqrt(cos(pi/4)) + 2*x
print(z)

```

3d)
```{r  eval=TRUE, include=TRUE}

# 3d)
round(z, 1)

```


## Concatenating multiple values in vectors

### Exercise 4

Start by defining the data vectors:

```{r eval=TRUE, include=TRUE}

# t.max vector
t.max <- c(20.7,18.9, 20.8, 18.8, 19.2, 18.6, 20.5, 
         20, 19.1, 21.3, 16.9, 18.6)

# est.clim vector
est.clim <- c("Anadia", "Dunas de Mira", "Nelas", "Guarda","Caramulo", "S. Jacinto", 
            "Viseu", "Serra da Muna", "Estarreja", "Fig. Cast. Rodrigo", "Arouca/S.Freita",
            "Moimenta da Beira")

```

4a) 
```{r eval=TRUE, include=TRUE}

# 4a)
sqrt(t.max + 100)

```

The function was applied to all elements - in R terminology this is a property of base functions called 'recycling'

4b)
```{r eval=TRUE, include=TRUE}

# 4b)
length(t.max)

```

4c)
```{r eval=TRUE, include=TRUE}

# 4c)
mean(t.max)
sd(t.max)

```

4d)
```{r eval=TRUE, include=TRUE}

# 4d)

# Calculate quartiles (the default behaviour of quantile function)
quantile(t.max)

# Make a boxplot
boxplot(t.max)

# Calculate the median
quantile(t.max)["50%"]

# or, more directly
median(t.max)

```


## Generating sequences of values


### Exercise 5


5a)
```{r eval=TRUE, include=TRUE}

# 5a)

k <- 10:100

sum(k)

```

5b) 
```{r eval=TRUE, include=TRUE}

# 5b)

# Make v vector
v <- seq(-pi, pi, by = pi/100)

# Compute the sin as u
u <- sin(v)

# Now let´s plot the two vectors in x and y
plot(v, u, type="l", lwd=2, col="blue", main="Sine function")

# Make the plot more pretty with some grid lines
abline(h=0,v=0,lty=2,col="black")
abline(v=pi/2,col="light grey",lty=2)
abline(v=-pi/2,col="light grey",lty=2)


# Alternatively, plotting a function can be accomplished more directly like this:
plot(sin, -pi, pi) # see ?plot.function

```


## Logical operations


### Quick exercises

QE 1)
```{r}
# correct answer:
print("c")

# Let's see one example of logical operations recycling:
a <- 1:20
a >= 10

```

QE 2)
```{r}
# correct answer:
print("c")

# Let's check
print(as.numeric(FALSE))
print(as.numeric(TRUE))

```

QE 3)
```{r}
# correct answer:
print("e")

# But let's check:
sum( (1:10) > 5 )

```


## Vector indexation


### Exercise 6 (logical operations and vector indexation)


6a)	
```{r eval=TRUE, include=TRUE}

# Input the data again (just in case... ;-)

t.max <- c(20.7, 18.9, 20.8, 18.8, 19.2, 18.6, 20.5, 20, 19.1, 21.3, 16.9, 18.6)

est.clim <- c("Anadia", "Mira", "Nelas", "Guarda", "Caramulo", "S. Jacinto", "Viseu", 
              "S. Muna", "Estarreja", "Fig. Cast. Rodrigo", "Arouca", "Moimenta")


## 6a) Now define the names attributed to each value in the t.max vector
names(t.max) <- est.clim
print(t.max)

```

6b)
```{r eval=TRUE, include=TRUE}

# 6b)
t.max[1:5]

```

6c)
```{r eval=TRUE, include=TRUE}

# 6c)
t.max[c("Anadia","Nelas")]

```

6d)
```{r eval=TRUE, include=TRUE}

# 6d) Combine two logical conditions separated by () and combined by the & (AND) operator
t.max[(t.max > 20) & (t.max <= 21)]

```

6e)
```{r eval=TRUE, include=TRUE}

# 6e)
# Notice that we accessing one element in the left-hand side and attributing a value 
# in the right-hand side
t.max["Guarda"] <- 19

# or using position indexation:
t.max[4] <- 19

```


# Matrix operations

---------------------------------------------------------------------------------------------------------


## Matrix indexing

### Exercise 7

7a)
```{r eval=TRUE, include=TRUE}

# 7a) The correct option is:
print("iv")

```

7b)
```{r eval=TRUE, include=TRUE}

# 7b) Input the data into R
smokers<-matrix(c(50.3, 60.6, 71.5, 82.3, 59.9, 79.3, 41.4, 80.9, 72.1, 59.1,
                  54.4, 52.4, 67.1, 78.3, 59.2, 65.1, 86.3, 81.3, 57.3, 61.3),
                ncol=10,nrow=2,byrow=TRUE)

# Add row and column names
rownames(smokers) <- c("W","M")
colnames(smokers) <- c("Aveiro", "Braga", "Bragança", "Porto", "Coimbra",
                       "Covilhã", "Leiria", "Lisboa", "Setúbal", "Faro")
print(smokers)

```
	
7c)
```{r eval=TRUE, include=TRUE}

# 7c) Index the column by name
mean(smokers["M",])
# or, by row number
mean(smokers[2,])

```
	
7d)
```{r eval=TRUE, include=TRUE}

# 7d) Use sorting based on the first row (women data by city)
#
sort(smokers[1,], decreasing=TRUE)[1:2]

```
	
7e)
```{r eval=TRUE, include=TRUE}

# 7e) Use sorting based on the absolute difference between groups
#
sort(abs(smokers[2,] - smokers[1,]), decreasing=TRUE)[1]

```

7f) 
```{r eval=TRUE, include=TRUE}

# 7f) calculate by column the index of where the maximum value is 1 - women and 2 - men
#
smkStats <- apply(smokers, 2, which.max)

# Now change the resulting vector from above
smkStats[smkStats == 1] <- "Women"
smkStats[smkStats == 2] <- "Men"

print(smkStats)

```



