---
title: 'R-intro - Session 4 (part 1 - dplyr) exercise solutions'
author: "João Gonçalves"
date: "31 de Julho de 2018"
output:
  pdf_document:
    toc: true
    toc_depth: 4
    fig_height: 0.5
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




## Objectives of session 4 (_dplyr_)

---------------------------------------------------------------------------------------------------------

* A brief tour to the `tidyverse`

* Data manipulation with the `dplyr` package
    + Main `dplyr` verbs
    + The pipe `%>%` operator


# The _dplyr_ package

---------------------------------------------------------------------------------------------------------

```{r, message=FALSE, warning=FALSE}

library(dplyr)
```


### Selecting columns using `select()`


#### QUICK EXERCISE 1

Using the `select` verb from the `dplyr` package and the `airquality` dataset, answer to the following questions:

1a.
```{r R.options=list(max.print=10)}

select(airquality, 1:3)
```

1b.
```{r R.options=list(max.print=10)}

select(airquality, -Ozone, -Temp)
```

1c.
```{r R.options=list(max.print=10)}

select(airquality, Solar.R:Day)
```



### Selecting rows with `filter()`

#### QUICK EXERCISE 2

2a.
```{r R.options=list(max.print=10)}

filter(iris, Petal.Length > 5)
```

2b.
```{r R.options=list(max.print=10)}

filter(iris, Petal.Length >= 3 & (Petal.Width < 1.2 | Sepal.Width < 2.8))
```

2c.
```{r R.options=list(max.print=10)}

filter(iris, Species == "versicolor")
# OR
filter(iris, Species %in% "versicolor")
```



### The pipe operator: `%>%`

### Sorting rows with `arrange()`

#### QUICK EXERCISE 3

Using the `arrange` verb from the `dplyr` package and the `iris` dataset, answer to the following question:

3a.
```{r R.options=list(max.print=10)}

iris %>% arrange(desc(Petal.Length), Sepal.Length)
```



### Adding new columns with `mutate()`


#### QUICK EXERCISE 4

4a.
```{r R.options=list(max.print=10)}

trees %>% mutate(Girth_cm = Girth * 2.54,
                 Height_mt = Height / 3.2808399)
```

4b.
```{r}

trees %>% 
  mutate(hvr = Height / Volume) %>%
  filter(hvr >= 3, hvr < 5) %>% 
  nrow
```



### Creating data summaries using `summarise()` and `group_by()`


#### QUICK EXERCISE 5

5a.
```{r}

PlantGrowth %>% 
  group_by(group) %>% 
  summarise(min = min(weight),
            max = max(weight),
            q25 = quantile(weight, probs=0.25),
            q50 = quantile(weight, probs=0.50),
            q75 = quantile(weight, probs=0.75),
            Nobs   = n())

```





